# Data visualisation {#datavis}

> Pre-requisite for this class: ch.3 ("Data visualisation") from R for Data Science - http://r4ds.had.co.nz/data-visualisation.html

At home you learned about the basic principles of data visualisation in R with the **ggplot2** package. Let us see how we can apply this to the Understanding Society data set.

Personally I can never remember all the details of the ggplot2 syntax. I often use the ready-made "recipes" from the R Graphics Cookbook by W.Chang -- https://www.amazon.co.uk/R-Graphics-Cookbook-Winston-Chang/dp/1449316956/. The 2nd edition is coming out later this year -- https://www.amazon.co.uk/Graphics-Cookbook-2e-Winston-Chang/dp/1491978600 .

You may also find Winston Chang's website useful (and not only for graphics) - http://www.cookbook-r.com .

## Reading in the data

First let us read in the data we used in week 2 when we learned about **dplyr** (a short version of the wave 1 data) and recreate the measures for weight, height and BMI.

```{r cache = TRUE}
library(tidyverse)
library(data.table)
W1 <- fread("exData/W1.csv")
W1 <- W1 %>%
  mutate(heightcm = ifelse(a_hlht == 1 & a_hlhtf > 0, 
                           a_hlhtf*30.48 + a_hlhti*2.54,
                           ifelse(a_hlht == 2 & a_hlhtc > 0, 
                                  a_hlhtc, NA))) %>%
  mutate(weightkg = ifelse(a_hlwt == 1 & a_hlwts > 0, 
                           a_hlwts*6.35 + a_hlwtp*0.45,
                           ifelse(a_hlwt == 2 & a_hlwtk > 0, 
                                  a_hlwtk, NA))) %>%
  mutate(bmi = weightkg / (heightcm / 100)^2)
head(W1, 3)
```

## Visualising one quantitative variable

**Exercise**. Visualise the distribution of the BMI with **ggplot2**. Which statistical graphs would be appropriate for this?


### Histogram.

```{r}
ggplot(W1, aes(x=bmi)) +
  geom_histogram(bins = 100) +
  xlab("Body mass index")
```

### Density chart.

```{r}
ggplot(W1, aes(x=bmi)) +
  geom_density() +
  xlab("Body mass index")
```


## Visualising one categorical variable

**Exercise**. Visualise the distribution of *a_ukborn* with **ggplot2**. Which statistical graphs would be appropriate for this?

### Bar plot.

```{r}
table(W1$a_ukborn)
W1 <- W1 %>%
  mutate(a_ukborn = ifelse(a_ukborn > 0, a_ukborn, NA)) %>%
  mutate(cbirth = recode(a_ukborn, "1" = "England", 
                         "2" = "Scotland",
                         "3" = "Wales",
                         "4" = "Northern Ireland",
                         "5" = "Not UK")) 
  
table(W1$cbirth)
W1 %>% 
  filter(!is.na(cbirth)) %>%
  ggplot(aes(x=cbirth)) +
  geom_bar() +
  xlab("Country of birth")

table(W1$cbirth, useNA = "always")
```

## Visualising two quantitative variables

**Exercise**. Visualise the joint distribution of weight (in kg) and height (in cm). In your chart show the regression line and the nonparametric smoothing line.

```{r}
ggplot(W1, aes(x = weightkg, y= heightcm)) +
  geom_point() +
  geom_smooth() +
  stat_smooth(method=lm)
```

## Visualising one categorical and one quantitative variable

**Exercise**. Visualise the distribution of BMI for a) men and women, b) different age groups.

## Visualising two categorical variables

**Exercise**. Use facets to visualise the distribution of *a_ukborn* by age group.

## Showing the relationships by group

**Exercise**. Use facets to visualise the association between age and BMI by ethnic group.


















