# Join data {#joining}

> For this class please read ch.13 on Relational Data from R for Data Science -- http://r4ds.had.co.nz/relational-data.html.

## Joining waves 1 and 2

I will start with joining data from waves 1 and 2 and then expand the algorithm to all the seven waves. First we need to read the data into R.

I am using the function **read_tsv()** from the **readr** package here since the data are tab separated.

```{r cache = TRUE}
# First I attach the packages I will use later. You need to install these packages first.
library(tidyverse)
library(data.table)
library(reshape2)

UndSoc1 <- read_tsv("data/UKDA-6614-tab/tab/us_w1/a_indresp.tab")
UndSoc2 <- read_tsv("data/UKDA-6614-tab/tab/us_w2/b_indresp.tab")

```

These are very large files that take a lot of space in the memory.

```{r cache = TRUE}
format(object.size(UndSoc1), units = "auto")
format(object.size(UndSoc2), units = "auto")

```

Before joining these two data frames I may want to select only the variables I need. These are:

* **pidp**: this is the unique cross-wave individual number (in both waves),

* **a_sex**: sex from wave 1,

* **a_dvage**: age from wave 1,

* **a_vote6**: level of interest in politics from wave 1,

* **b_sex**: sex from wave 2,

* **b_dvage**: age from wave 2,

* **b_vote6**: level of interest in politics from wave 2.

Note that for now I keep it intentionally simple.

First I will edit both data sets to keep only the variables I need.

```{r cache = TRUE}
UndSoc1ed <- UndSoc1 %>%
        select(pidp, a_sex, a_dvage, a_vote6)
UndSoc2ed <- UndSoc2 %>%
        select(pidp, b_sex, b_dvage, b_vote6)

```

Note that these are much smaller objects.

```{r cache = TRUE}
format(object.size(UndSoc1ed), units = "auto")
format(object.size(UndSoc2ed), units = "auto")

```

I then remove the larger data sets from the memory to free it up.

```{r cache = TRUE}
rm(UndSoc1, UndSoc2)
```

Let us explore the data.

```{r cache = TRUE}
head(UndSoc1ed)
head(UndSoc2ed)

```

Now we can join the data sets. As you know already, there can be several types of joins. Inner join will keep the observations that match in both data frames.

```{r cache = TRUE}
inner <- UndSoc1ed %>%
        inner_join(UndSoc2ed, by = "pidp")
```

Note that I use the variable **pidp** as the key for joining.

The same result can be achieved with the function **merge** from base R.

```{r cache = TRUE}
inner2 <- UndSoc1ed %>%
        merge(UndSoc2ed, by = "pidp")
identical(as.data.frame(inner), inner2)
```

These two data frames are identical. Note the use of **as.data.frame()**. **inner** is a tibble and unless converted into a simple data frame it is not identical to **inner2** because of the differences in object type.

Inner join will only retain individuals that are present in both waves. This is why we only have `r nrow(inner)` observations in the joined data compared to `r nrow(UndSoc1ed)` in wave 1. But do we want this? Imagine that the person is present in wave 1, does not participate in wave 2, but then re-appears in wave 3. We probably want to include such individuals as well.

Left join will keep all individuals in wave 1 and only those in wave 2 that can be matched to them.

```{r cache = TRUE}
left <- UndSoc1ed %>%
        left_join(UndSoc2ed, by = "pidp")
# or, identically,
# left <- UndSoc1ed %>%
#         merge(UndSoc2ed, by = "pidp", all.x = TRUE)
```

You can check that the number of individuals in wave 1 and the joined data frame is the same.

Right join will keep all individuals from wave 2 and only those from wave 1 that can be matched to them.

```{r cache = TRUE}
right <- UndSoc1ed %>%
        right_join(UndSoc2ed, by = "pidp")
# or, identically,
# right <- UndSoc1ed %>%
#         merge(UndSoc2ed, by = "pidp", all.y = TRUE)
```

Usually I would want all the individuals from both waves to remain in the data set, no matter if they can be matched to other waves. If I need to exclude them from the analysis I prefer to do this manually. This can be achieved with full join.

```{r cache = TRUE}
full <- UndSoc1ed %>%
        full_join(UndSoc2ed, by = "pidp")
# or, identically,
# full <- UndSoc1ed %>%
#          merge(UndSoc2ed, by = "pidp", all = TRUE)
```

Note that **full** has `r nrow(full)` observations compared to  `r nrow(UndSoc1ed)` in wave 1 and `r nrow(UndSoc2ed)` in wave 2. This is because it includes a) all individuals that took part in both waves 1 and 2, b) those who took part in wave 1, but not in wave 2, c) those who missed wave 1 but joined the study in wave 2.

Let us explore the joined data set.

```{r cache = TRUE}
head(full)
```

We have missing values for wave 1 variable for those people who joined in wave 2.
