# Read data {#readdata}

> Please read ch.11 (Data import) from R for Data Science - http://r4ds.had.co.nz/data-import.html

For this module we will use data from the Understanding Society survey (https://discover.ukdataservice.ac.uk/catalogue/?sn=6614). I assume that you have famliarised yourself with the data set, registered an account with the UK Data Service website and downloaded the data in the tab format.

The first thing we need to do is to read the data in R. There are multiple ways of doing this.

For all the examples I will use the individual adult data from wave 1 (*a_indresp.tab*).

## Base R

In base R we have the **read.table** function. I will wrap it into the **system.time** function to measure how long the execution will take.

```{r}
system.time(UndSoc1 <- read.table("data/UKDA-6614-tab/tab/us_w1/a_indresp.tab",
                      header = TRUE,
                      stringsAsFactors = FALSE))
```

I set *header = TRUE* to make sure that the first row in the data is interpreted as variable names. *stringsAsFacrors = FALSE* means that the text variables will be read in as character vectors rather than factors. We can convert them into factors when necessary.

## Package readr

We can also read in these data with the package **readr** (part of **tidyverse**). The main advantage is that it works faster.

```{r}
library(readr)
system.time(UndSoc2 <- read_tsv("data/UKDA-6614-tab/tab/us_w1/a_indresp.tab"))
```

**readr** was able to read the data set much faster than base R.

## Package data.table

The **fread** function from the **data.table** package is probably the fastest way to read in the data. 

```{r}
library(data.table)
system.time(UndSoc3 <- fread("data/UKDA-6614-tab/tab/us_w1/a_indresp.tab"))
```

It took less than 3 seconds!
 
With small data sets the choice between these three methods is not very important, but with larger data the gain in efficiency that **readr** and **data.table** provide can be quite substantial.

## Other data formats

In R you can easily read in data in other formats, such as csv files, Stata, SPSS, SAS, Excel and others. There are many tutorials on how to do this on the web. See, for example, https://www.datacamp.com/courses/importing-data-in-r-part-1/ and https://www.datacamp.com/courses/importing-data-in-r-part-2 .

## Saving the R workspace

Once you have read your data into R you can save it as R workspace.

```{r}
# I will remove some objects from memory to speed things up
rm(UndSoc2, UndSoc3)
# saving R workspavce now in myData (you need to create myData first)
save.image("myData/readTest.RData")
```

Next time I need this file I can simply load the workspace.

```{r}
# first let's remove everything from the workspace
rm(list = ls())
# load the workspace
system.time(load("myData/readTest.RData"))
```

Of course, in R workspace you can save not only data frames but any objects: models, plots, functions, etc.



 



