# Tidy data

In the previous part of the course (joiningData.Rmd) we learned how to join together data from seven waves of the Understanding Society. Let us open this data set.

```{r}
UndSoc <- read_tsv("myData/all7.tab")
head(UndSoc)
```

Now we will work on how these data can be represented and prepared for the analysis. Please read ch.12 on Tidy Data from the R for Data Science Book -- http://r4ds.had.co.nz/tidy-data.html.

## Long and wide formats

Let us keep only a few observations and columns in the data and more closely look at its structure.

```{r}
UndSocExample <- UndSoc %>%
  filter(pidp == 68001367 | pidp == 68004087) %>%
  select(pidp, a_sex: b_vote6)
UndSocExample
```

These are the data for two individuals only in waves 1 and 2. The data are represented in the wide format. This means that we have one row for each individual, and data from different waves are recorded in several columns. For example, the data on sex from wave 1 is in column *a_sex* and the data on sex from wave is in *b_sex*.

You will find this representation of the data common in longitudinal data sets. It may be convenient for certain purposes, but it is generally recommended to keep the data in the long format (that corresponds to the tidy data principles as described in the R for Data Science book).

To move from the wide to the long format we can use the function **melt** and **cast** functions from the **reshape2** package.

```{r}
require(reshape2)

# First we "melt" the data frame.
UndSocExampleMolten <- UndSocExample %>%
  melt(id = "pidp")
UndSocExampleMolten

# Next I want to split the column variable into a column indicating wave and a column indicating variable name. 
# I will use the function separate() from tidyr.

UndSocExampleSep <- UndSocExampleMolten %>%
        separate(variable, into = c("wave", "variable"), sep = "_")
UndSocExampleSep

# Next we "cast" the molten data frame into the format we want.

UndSocExampleLong <- UndSocExampleSep %>%
  dcast(pidp + wave ~ variable)
UndSocExampleLong
```

Now the data are in the "long format". This means that we have as many rows for each individual as the number of waves, a variable indicating wave, and all other variables are in columns. In most cases with longitudinal data, the long format is easier to work with.

What if we want to convert the data back to the wide format?

```{r}
# First melt
UndSocExampleMolten2 <- UndSocExampleLong %>%
  melt(id = c("pidp", "wave"))
UndSocExampleMolten2

# Unite the columns
UndSocExampleUnited <- UndSocExampleMolten2 %>%
  unite("variable", c("wave", "variable"), sep = "_")
UndSocExampleUnited

# And now cast
UndSocExampleWide <- UndSocExampleUnited %>%
  dcast(pidp ~ variable)
UndSocExampleWide
```

We can also restructure the data using the **gather** and **spread** functions from the **tidyr** package (part of **tidyverse**). **gather** is roughy equivalent to **melt** and **spread** is roughy equivalent to **dcast**.

Moving from wide to long:

```{r}
UndSocExample

# This "melts" the data frame. 
UndSocExample %>%
  gather(a_sex:b_vote6, key = "variable", value = "value")

# Next we want to split the "variable" column and "cast" in the long format
UndSocExample %>%
  gather(a_sex:b_vote6, key = "variable", value = "value") %>%
  separate(variable, into = c("wave", "variable"), sep = "_") %>%
  spread(key = variable, value = value)

```

If we want to move from long to wide:

```{r}
UndSocExampleLong

UndSocExampleLong %>%
  gather(dvage:vote6, key = "variable", value = "value") %>%
  unite("variable", c("wave", "variable"), sep = "_") %>%
  spread(key = variable, value = value)

```

**Exercise**. Reshape the full **UndSoc** data frame from wide to long format. Call the object where you will
store the result **UndSocLong**.

```{r}
UndSocLong <- UndSoc %>%
  gather(a_sex:g_vote6, key = "variable", value = "value") %>%
  separate(variable, into = c("wave", "variable"), sep = "_") %>%
  spread(key = variable, value = value)
UndSocLong
```

## Cleaning the data

Before we begin the analysis we want to make sure that the data have been cleaned and all the missing values have been correctly identified. It usually makes sense to separate the cleaning and analysis stages into separate scripts.

Let us explore the data set we have. Note that if we had not converted the data into the long format we would have to tabulate and clean each variable seven times.

```{r}
summary(UndSocLong)
table(UndSocLong$wave)
table(UndSocLong$dvage)
table(UndSocLong$sex)
table(UndSocLong$vote6)
```

Note the negative values for dvage, sex and vote6. These are missing values that need to be coded as missing.

```{r}
UndSocLong  <- UndSocLong %>%
  mutate(dvage = ifelse(dvage > 0, dvage, NA)) %>%
  mutate(sex = ifelse(sex > 0, sex, NA)) %>%
  mutate(vote6 = ifelse(vote6 > 0, vote6, NA))
table(UndSocLong$dvage)
table(UndSocLong$sex)
table(UndSocLong$vote6)
```

We may also want to code **sex** as "male" and "female" and assign meaningful labels to **vote6** converting these variables into factors.

```{r}
UndSocLongClean <- UndSocLong %>%
  mutate(sex = recode(sex, "1" = "male", "2" = "female")) %>%
  mutate(vote6 = recode(vote6, "1" = "very", "2" = "fairly", "3" = "not very", "4" = "not al all")) %>%
  mutate(sex = factor(sex)) %>%
  mutate(vote6 = factor(vote6))
UndSocLongClean

write_tsv(UndSocLongClean, "myData/all7clean.tab")

```

## Homework

To prepare for the class on 19 February please read ch. 3 (Data visualisation) from the R for Data Science book and do all the exercises -- http://r4ds.had.co.nz/data-visualisation.html

Please also read ch.12 on tidy data and do all the exercises -- http://r4ds.had.co.nz/tidy-data.html



